Modularly Composing Typed Language Fragments

### 2014-11-02 att-pldi15/att-pldi15.tex
More edits

In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant on record types and constrained string types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of its associated term-level operators (e.g. record projection) using functions written in a static language where types and translations are values. The semantics provide  strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that all \emph{tycon  invariants} will be conserved under extension. Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing barriers around tycons using type abstraction, the same principle that underlies reasoning with ML-style modules.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation
\section{Introduction}\label{intro}
Typed programming languages are most often described as being composed of \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes fragments around type constructors, describing each in a different chapter \cite{pfpl}. Languages are then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that includes partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
Another related approach is to describe fragments with a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.

@@

Universality properties (e.g. ``Turing-completeness'') can usually be invoked to guarantee that an embedding that preserves a desirable fragment's dynamics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a fragment like this is needed, a new \emph{dialect} of a language must be constructed. Within the ML lineage, for example, dialects abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

@@

These problems do not arise for a fragment that can be expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

@@

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons control the semantics of their associated term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not necessarily provide mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about language fragments. We conclude by discussing  related and future work in Sec. \ref{prior-work}.

@@

This basic separation of the EL and IL is commonly used for full-scale language specifications, e.g. the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}. The internal language is purposely kept small, e.g. defining only simple products, to simplify metatheoretic reasoning and compilation. The EL then specifies various useful higher-level constructs, e.g. record types, by translation to the IL. In @$\lambda$, the EL builds in only function types. All other external constructs are defined in the \emph{tycon context}, described in the Sec. \ref{types}.

@@

 External typing contexts $\Upsilon$ map variables to types, so the type translation  judgement induces judgement  $\vdash_\Phi \Upsilon \leadsto \Gamma$.

@@

We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}. 
%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
In examples, we will omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences (e.g. writing $\Gamma_\text{test} := x : \tau$).

The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}. The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$. Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and key metatheoretic properties.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq. The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.  Only three new  kinds are needed for the SL to serve its role as the language used to  statically compute over types and type and term translations: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

@@

All user-defined tycons must be defined in the \emph{tycon context}, $\Phi$, which is simply a list of tycon definitions. Each tycon defines a name, $\tc$, a \emph{tycon structure}, $\theta$, and a \emph{tycon signature}, $\psi$. We will return to the tycon structure below. Tycon signatures have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}, which we   discuss in Sec. \ref{external-terms}. The first premise of \rulename{k-ty} extracts the index kind and the second checks the type index against it.

@@

\paragraph{Examples}
Our first example of a user-defined tycon is $\tcvar{rstr}$. It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way). Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

@@

\subsection{Type Translations}\label{sec:type-translations}
Recall that every type $\st$ must have a type translation, $\tau$. Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$. A tycon with index kind $\ktyidx$ defines a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).

The kind $\kity$ has a single introductory form, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}. Each form in the syntax for internal types, $\tau$,  corresponds to a form in the syntax of translational internal types, $\qity$. For example, our translation schema for $\tcvar{rstr}$  simply chooses to ignore the type index and represent all regular strings internally as strings, of internal type abbreviated $\keyw{str}$. We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define the translation schema as $\small\theta_\text{rstr} := \tcstruct{\slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}}{\omega_\text{rstr}}$.

@@

We choose a translation schema for $\tcvar{lprod}$ that generates nested binary product types by recursing over the type index and referring to the translations of the types therein (though this is not the only workable choice, e.g. we could also have used a list). We assume $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}~\keyw{in}~\omega_\text{lprod}$  where $\stx{lprod/trans}:=$
\[\small
\begin{array}{l}
\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\skap{\kity}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~\sqity{\iunit}~\\
  \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
\end{array}\]
Evaluating this translation schema with the index of $\stx{paper}$, for example, produces the value $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$. Note that we do not include logic to optimize away the trailing unit type for simplicity (and to again emphasize that many translations are possible for any given type).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away

\subsubsection{Selective Type Translation Abstraction}\label{sec:selective-type-translation-abstraction}
References to translations of other types are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to our main results. This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking. The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the ``delegated tycon'', $c$, are replaced by a corresponding type variables, $\alpha$. The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the corresponding between types, their actual translations and the unique type variables which appear in their place. For example, $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
where $\small\tau_\text{paper/abs} := \alpha_1 \times (\alpha_2 \times \iunit)$ and 
$\memD_\text{paper/abs}  := \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2$.

Each type translation store induces a \emph{type substitution}, $\delta$, and a corresponding internal type formation context, $\Delta$, according to the judgement $\memD \leadsto \delta : \Delta$. Type substitutions simply define an $n$-ary substitution for type variables, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$. For example, $\memD_\text{paper/abs} \leadsto \delta_\text{paper/abs} : \Delta_\text{paper/abs}$ where $\delta_\text{paper/abs} := \keyw{str}/\alpha_1, \keyw{str}/\alpha_2$ and $\Delta_\text{paper/abs} := \alpha_1, \alpha_2$. 
We can apply type substitutions to internal types, terms and typing contexts, written $[\delta]\ity$, $[\delta]\iota$ and $[\delta]\Gamma$, respectively. For example, $[\delta_\text{paper/abs}]\tau_\text{paper/abs}$ is $\tau_\text{paper} := \keyw{str} \times (\keyw{str} \times \iunit)$, i.e. the actual type translation of $\stx{paper}$. Indeed, we can now give the rule defining the type translation judgement, $\vdash_\Phi \st \leadsto \tau$, mentioned in Sec. \ref{overview}. We simply determine any selectively abstract translation, then apply the induced substitution:
\begin{mathpar}\small
\inferrule[conc-ty-trans]{
    \istype{\st}{\Phi}\\
    \tdeabs{\Phi}{\tc}{\srep{\st}}{\emptyset}{\tau}{\memD}\\
    \memD \leadsto \delta : \Delta
}{
    \vdash_\Phi \st \leadsto [\delta]\tau
}
\end{mathpar}

@@

Having established how types are constructed, and how they determine selectively abstract and from there actual type translations, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.

@@

The fourth premise extracts the \emph{intro opcon definition} from the \emph{opcon structure}, $\omega$, of the tycon structure, calling it $\stx{def}$. This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so. The function has access to the type index, the term index and an interface to the list of arguments, and its kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.

@@

The rest of this opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces. We assume a straightforward helper function, $\svar{listfold3}$, that raises an error if the three lists are not of the same length. The base case is the translational empty product. The recursive case first checks that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$. Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$. Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $\st$ is the value of $\svar{rowty}$, if the argument is well-typed, and raises an error if not. The final line constructs a nested tuple based on this translation and the recursive result. Taken together, the translational internal term that will be generated for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$.

@@

The argument interface list generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$, is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.

@@

\newpage
% We recommend abbrvnat bibliography style.

>>>>

In this paper, we begin from first principles by specifying a language, @$\lambda$, in the style of many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant on record types and constrained string types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of its  term-level operators (e.g. record projection) using a static language where types and translations are values. The semantics includes a simple check that leads to powerful metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that  \emph{tycon invariants} will always be conserved under extension. Type system providers need not provide mechanized proofs. Instead, the check enforces a \emph{translation independence} property via type abstraction, the same principle underlying ML-style module systems.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation
\section{Introduction}\label{intro}
Typed programming languages are most often described as being composed from \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes such fragments around type constructors, describing each in a different chapter \cite{pfpl}. Languages are then identified by a set of these type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
In the literature, researchers often describe fragments using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.

@@

Universality properties (e.g. ``Turing-completeness'') are often invoked to guarantee that an embedding that preserves a desirable fragment's dynamics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a fragment must be exposed directly, rather than as a library, a new \emph{dialect} of a language must be constructed. Within the ML lineage, for example, dialects that go beyond ``core ML'' abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

@@

These problems do not arise for a fragment that can be expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client. Other langauges offer similar facilities, sometimes called \emph{abstract data types}.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

@@

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons control the semantics of their term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not provide mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We conclude by discussing  related and future work in Sec. \ref{prior-work}.

@@

This basic separation of the EL and IL is commonly used for full-scale language specifications, e.g. the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}. The internal language is purposely kept small, e.g. defining only binary products, to simplify metatheoretic reasoning and compilation. The EL then specifies various useful higher-level constructs, e.g. record types, by translation to the IL. In @$\lambda$, the EL builds in only function types. All other external constructs are defined in the \emph{tycon context}, described starting in the Sec. \ref{types}.

@@

External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$, where $\Gamma$ maps variables to the corresponding type translation. In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.

@@

We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}. 
%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}. The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$. Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq. The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.  Only three new  kinds are needed for the SL to serve its role as the language used to   control typechecking and translation of the EL: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

@@

All user-defined tycons $\tc$ must be defined in the \emph{tycon context}, $\Phi$, which is simply a list of tycon definitions, $\tcdef{\tc}{\psi}{\theta}$, where $\theta$ is the \emph{tycon structure} and $\psi$ is the \emph{tycon signature}. We will return to tycon structures below. Tycon signatures have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}, which we   discuss in Sec. \ref{external-terms}. The first premise of \rulename{k-ty} extracts the index kind and the second checks the type index against it.

@@

\paragraph{Examples}
Our first example of a user-defined tycon is $\tcvar{rstr}$. It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way). Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which statically obey the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

@@

\subsection{Type Translations}\label{sec:type-translations}
Recall that every type $\st$ must have a type translation, $\tau$. Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$. A tycon with index kind $\ktyidx$ must define a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).

The kind $\kity$ has a single introductory form, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}. Each form in the syntax for internal types, $\tau$,  corresponds to a form in the syntax of translational internal types, $\qity$. For example, our schema for $\tcvar{rstr}$ will simply choose to ignore the type index and translate regular strings to strings, of internal type abbreviated $\keyw{str}$. We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define $\small\theta_\text{rstr} := \tcstruct{\slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}}{\omega_\text{rstr}}$.

@@

These are used in the schema for $\tcvar{lprod}$, which generates nested binary product types by recursing over the type index and referring to the translations of the types therein (this is not the only viable choice, e.g. we could also have used a list). We assume $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}~\keyw{in}~\omega_\text{lprod}$  where $\stx{lprod/trans}:=$
\[\small
\begin{array}{l}
\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\skap{\kity}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~\sqity{\iunit}~\\
  \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
\end{array}\]
Evaluating this translation schema with the index of $\stx{paper}$, for example, produces the value $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$. Note that we do not include logic to optimize away the trailing unit type for simplicity (and to again emphasize that many translations are possible for any given type).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away

\subsubsection{Selective Type Translation Abstraction}\label{sec:selective-type-translation-abstraction}
References to type translations are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to our main results. This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking. The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the ``delegated tycon'', $c$, are replaced by a corresponding type variable, $\alpha$. The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the correspondence between types, their actual translations and the unique type variables which appear in their place, e.g. $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
where $\small\tau_\text{paper/abs} := \alpha_1 \times (\alpha_2 \times \iunit)$ and 
$\memD_\text{paper/abs}  := \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2$.

Each type translation store induces a \emph{type substitution}, $\delta$, and a corresponding internal type formation context, $\Delta$, according to the judgement $\memD \leadsto \delta : \Delta$. Type substitutions simply define an $n$-ary substitution for type variables, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$. For example, $\memD_\text{paper/abs} \leadsto \delta_\text{paper/abs} : \Delta_\text{paper/abs}$ where $\delta_\text{paper/abs} := \keyw{str}/\alpha_1, \keyw{str}/\alpha_2$ and $\Delta_\text{paper/abs} := \alpha_1, \alpha_2$. 
We can apply type substitutions to internal types, terms and typing contexts, written $[\delta]\ity$, $[\delta]\iota$ and $[\delta]\Gamma$, respectively. For example, $[\delta_\text{paper/abs}]\tau_\text{paper/abs}$ is $\tau_\text{paper} := \keyw{str} \times (\keyw{str} \times \iunit)$, i.e. the actual type translation of $\stx{paper}$. Indeed, we can now give the rule defining the type translation judgement, $\vdash_\Phi \st \leadsto \tau$, mentioned in Sec. \ref{overview}. We simply determine any selectively abstract translation, then apply the induced substitution:
\begin{mathpar}\small
\inferrule[ty-trans]{
    \istype{\st}{\Phi}\\
    \tdeabs{\Phi}{\tc}{\srep{\st}}{\emptyset}{\tau}{\memD}\\
    \memD \leadsto \delta : \Delta
}{
    \vdash_\Phi \st \leadsto [\delta]\tau
}
\end{mathpar}

@@

Having established how types are constructed, and how they determine selectively abstracted and from there actual type translations, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.

@@

The fourth premise extracts the \emph{intro opcon definition} from the \emph{opcon structure}, $\omega$, of the tycon structure, calling it $\stx{def}$. This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so. The function has access to the type index, the term index and an interface to the list of arguments, and is kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.

@@

The rest of this opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces. We assume a straightforward helper function, $\svar{listrec3}$, that raises an error if the three lists are not of the same length. The base case is the translational empty product. The recursive case first checks that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$. Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$. Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $n$ is the position of $\svar{rowarg}$ in $\svar{args}$ and $\st$ is the value of $\svar{rowty}$, if the argument is well-typed, and raises an error if not. The final line constructs a nested tuple based on this translation and the recursive result. Taken together, the translational internal term that will be generated for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$.

@@

The argument interface list is is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$. It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$.

@@

\newpage~\newpage
% We recommend abbrvnat bibliography style.
